\documentclass[12pt]{article}
\usepackage{fullpage}
\usepackage{titlesec}
\usepackage{tikz}
\usepackage{amsfonts,amssymb}
\usepackage{amsmath}
\usepackage{comment}
\usetikzlibrary{automata, positioning}

\input ../libraries/mac.tex
\input ../libraries/mathmac.tex

\begin{document}
\pagestyle{plain}
\titleformat{\subsection}[runin]
  {\normalfont\large\bfseries}{\thesubsection}{1em}{}

\title{Homework 4}
\author{Brooke Fugate, Michael O'Connor, Rohan Shah}
\date{}

\maketitle

\section*{Problem B7}
\subsection*{(a)} S is the starting symbol.
\newline
$S \rightarrow c$ \newline
$S \rightarrow aSa$ \newline
$S \rightarrow bSb$ \newline
The smallest possible string is c which is given by the first production.
We can build from the beginning and end of the string simultaneously with the
other two productions to ensure that the string following the c is the reverse
of the string before the c.

\subsection*{(b)} S is the starting symbol.
\newline
$S \rightarrow T$ \newline
$T \rightarrow aTBb$ \newline
$B \rightarrow b$ \newline
$B \rightarrow \epsilon$ \newline
$T \rightarrow ab$ \newline
$T \rightarrow abb$ \newline
The smallest possible strings are ab/abb which is given by replacing S with T
and T with ab or abb. We build up the string with by replacting T with aTBb.
That guarantees that $n \ge m$ and since B can either be b or $\epsilon$
$n \le 2m$. For $K \ge 2$, same as above, except with $T \rightarrow aTBb$
changed to $T \rightarrow aTB_1B_2...B_Kb$, where the productions for $B_i$ are
$B_i \rightarrow b$ and $B_i \rightarrow \epsilon$ and we have the productions
$T \rightarrow abb, T \rightarrow abbb ... T \rightarrow abbb...b$ with the
number of $b$'s in the last production equal to the value of K. The smallest
possible string is ab which is given by replacing S with T and T with ab.
We build up the string with by replacting T with aTBb. That guarantees that
$n > m$ and since any $B_j \mid 1 \le j \le k$ can either be b or
$\epsilon$ $n \le km$.

\subsection*{(c)} S is the starting symbol.
\newline
$S \rightarrow S_1$ \newline
$S \rightarrow S_2$ \newline
$S_1 \rightarrow aCb$ \newline
$C \rightarrow aCb$ \newline
$C \rightarrow \epsilon $ \newline
$S_2 \rightarrow aBbb$ \newline
$B \rightarrow aBbb$ \newline
$B \rightarrow \epsilon $ \newline
We create the union of the two languages using the productions
$S \rightarrow S_1$ and $S \rightarrow S_2$. The non-terminal $S_1$ is then the
starting symbol for the language $\{a^{n}b^{n}\mid n \geq 1\}$, which has a
smallest string of $ab$ by the production $S_1 \rightarrow aCb$ and can grow
only with the production $C \rightarrow aCb$ when an $a$ and $b$ are added at
the same time. The non-terminal $S_1$ is then the starting symbol for the
language $\{a^{n}b^{2n}\mid n \geq 1\}$, which has a smallest string of $abb$
by the production $S_2 \rightarrow aBbb$ and can grow only by the production
$B \rightarrow aBbb$ when an $a$ and two $b$'s are added at the same time.

\subsection*{(d)} S is the starting symbol.
\newline
$S \rightarrow S_1$ \newline
$S \rightarrow S_2$ \newline
$S_1 \rightarrow aAaBb$ \newline
$A \rightarrow aAa$ \newline
$A \rightarrow Bb$ \newline
$B \rightarrow Bb$ \newline
$B \rightarrow \epsilon $ \newline
$S_2 \rightarrow aAbbbbBbbbb$ \newline
$A \rightarrow aA$ \newline
$A \rightarrow \epsilon$ \newline
$B \rightarrow bbbbaBbbbb$ \newline
$B \rightarrow a $ \newline
$S_1$ produces the language $a^mb^na^mb^p$, the smallest string S1 can produce
is abab, by producing aAaBb from $S_1$ then producing Bb from A and then
$\epsilon$ from B and then again producing $\epsilon$ from the remaining B.
Instead of directly producing a single b from A we could instead produce
more aAa constructions making m arbritrarily large, then finally producing an
arbritary amount of b by producing Bb from A and then more Bb from B.
We can do the same for the last B to get another arbrirtrary amount of b,
having $m, n, \text{ and } p \ge 1$.\newline $S_2$ produces the language
$a^mb^{4n}a^pb^{4n}$, the smallest string that $S_2$ produces is abbbbabbbb
which is produced by replacing A with $\epsilon$ and B with a.
We can then create an arbitrary $a^m$ by producing aA from A, we also can
produce an arbritrary $a^p$ by producing bbbbaBbbbb from B, which will also
produce an arbritrary $b^{4n}$.\newline S can go to either $S_1$ or $S_2$ so
the language is $a^mb^na^mb^p \cup a^mb^{4n}a^pb^{4n}$.

\subsection*{(e)} S is the starting symbol.
\newline
$S \rightarrow C$ \newline
$C \rightarrow ACB$ \newline
$A \rightarrow aa$ \newline
$A \rightarrow ab$ \newline
$A \rightarrow ba$ \newline
$A \rightarrow bb$ \newline
$B \rightarrow a$ \newline
$B \rightarrow b$ \newline
$C \rightarrow \epsilon$ \newline
The smallest string that S creates is $xy \mid x = \text{ combination of 2 }
a/b \text{ characters}, y =  a \text{ or } b$, so $|x| = 2|y|$, we get this by
producing ACB from C and then a final production from A and B.  We could keep
producing ACB from C to get an arbritarily large $|y|$ and therefore
$|x| = 2|y|$.

\end{document}
