\documentclass[12pt]{article}
\usepackage{fullpage}
\usepackage{titlesec}
\usepackage{tikz}
\usepackage{amsfonts,amssymb}
\usepackage{amsmath}
\usepackage{comment}
\usetikzlibrary{automata, positioning}

\input ../libraries/mac.tex
\input ../libraries/mathmac.tex

\begin{document}
\pagestyle{plain}
\titleformat{\subsection}[runin]
  {\normalfont\large\bfseries}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[runin]
  {\bfseries}{}{1em}{}

\title{Homework 6}
\author{Brooke Fugate, Michael O'Connor, Rohan Shah}
\date{}

\maketitle

\section*{Problem B5}
\subsection*{a}
Prove that telling if a partial recursive function is a constant function is undecidable.  The property that a partially recursive function is a constant function, denotated P, is decidable iff the set $P_C = \{P(a)|a \in C and P(a)\}$ is recursive, where C is the set of partially recursive functions that are constant functions. By the Rice Theorem, the set $P_C$ is non-recursive unless C is trivial.  For C to be trivial it must either be the empty set or be the set of all recursive functions. The function $f(x)=0$ is partially recursive and is a constant set, so C is not empty.  The successor function is an example of a partially recursive function that is not constant, so C is not the set of all partially recursive functions.  So we know that C is not trivial, therefore $P_C$ is not recursive and P must be undecidable. 
\subsection*{b} 
Prove that the property of two partial recursive functions being identical is undecidable.  We will reduce the set $P_B = \{x,y \in \mathbb{N} | \varphi_x = \varphi_y\}$ to the set $P_C = \{x \in \mathbb{N} | \varphi_x = \varphi_a \}$.  Since we know C is undecidable from part c) reducing C to B will show that B is undecidable. To show that C is reducable to B  we need to show that there is a function from $f : C \rightarrow B$.  $x \in C \text{ iff } f(x) \in B$ so if $f(x) = < x , a >$ then C is reducable to B, therefore B is undecidable.
\subsection*{c} 
Prove that the property of a partial recursive function $\varphi_x$ is equal to a given partial recursive function $\varphi_a$ is not decidable.  As in (a), we must show that the set $P_C = \{x \in \mathbb{N} | \varphi_x = \varphi_a \}$ to is non-trivial.  The set is not empty because $\varphi_a = \varphi_a$ and the set is not the set of all partially recursive functions because $S (\varphi_a(x)) \neq \varphi_a(x)$, so the set is not trivial and is therefore non-recursive, and the property on partially recursive functions is undecidable.  
\subsection*{d}
Prove that telling if a partial recursive function diverges for all input is not decidable.  As in (c), we must show that the set $P_C = \{x \in \mathbb{N} | \varphi_x \text{is undecidable for all inputs} \}$ is non-trivial.  The set is not empty because we can define a function $\forall x \mid \varphi_a(x) = undecidable$ and the set is not the set of all partially recursive functions because the successor function is not undecidable on all inputs, so the set is not trivial and is therefore non-recursive, and the property on partially recursive functions is undecidable.  
\end{document}
