\documentclass[12pt]{article}
\usepackage{amsfonts,amssymb}
\usepackage{amsmath}

\setlength{\topmargin}{-.5in}
\setlength{\oddsidemargin}{0 in}
\setlength{\evensidemargin}{0 in}
\setlength{\textwidth}{6.5truein}
\setlength{\textheight}{8.5truein}

\input ../libraries/mac.tex
\input ../libraries/mathmac.tex

\input xy
\xyoption{all}
\def\fseq#1#2{(#1_{#2})_{#2\geq 1}}
\def\fsseq#1#2#3{(#1_{#3(#2)})_{#2\geq 1}}
\def\qleq{\sqsubseteq}

\begin{document}
\begin{center}
\fbox{{\Large\bf Spring, 2014 \hspace*{0.4cm} CIS 511}}\\
\vspace{1cm}
{\Large\bf Introduction to the Theory of Computation\\
Jean Gallier \\
\vspace{0.5cm}
Homework 1}\\[10pt]
January 21, 2014; Due February 4, 2014, {\it beginning of class}\\
\end{center}

``A problems'' are for practice only, and should not
be turned in.

\medskip\noindent
{\bf Problem A1.} 
Given an alphabet $\Sigma$, prove
that the relation $\leq_1$ over $\Sigma^*$ defined such that
$u \leq_1 v$ iff $u$ is a prefix of $v$, is a partial ordering.
Prove that the relation $\leq_2$ over $\Sigma^*$ defined such that
$u \leq_2 v$ iff $u$ is a substring of $v$, is a partial ordering.


\medskip\noindent
{\bf Problem A2.} 
Given an alphabet $\Sigma$, for any language $L\subseteq \Sigma^*$,
prove that $L^{**} = L^*$ and $L^*L^* = L^*$.

\vspace {0.5cm}\noindent
{\bf Problem A3.} 
Let $D = (Q,\Sigma,\delta,q_0,F)$ be a DFA.
Prove that for all $p\in Q$ and all $u, v\in \Sigma^*$,
\[
\delta^*(p, uv) = \delta^*(\delta^*(p, u), v).
\]

\vspace {0.5cm}
``B problems'' must be turned in.

\vspace {0.25cm}\noindent
{\bf Problem B1 (30 pts).} 
Let $D = (Q,\Sigma,\delta,q_0,F)$ be a DFA.
Recall that a state $p\in Q$ is {\it accessible\/} or {\it reachable\/}
iff there is some string $w\in\Sigma^*$, such that
$$\delta^{*}(q_0,w) = p,$$
i.e., there is some path from $q_0$ to $p$ in $D$.
Consider the following method for computing the set $Q_{r}$ of
reachable states (of $D$): define the sequence of sets
$Q_{r}^{i}\subseteq Q$, where

\medskip
$Q_{r}^{0} = \{q_0\}$,
 
\medskip
$Q_{r}^{i+1} = \{q\in Q \mid \exists p\in Q_{r}^{i}, \exists a\in\Sigma,\>
q = \delta(p,a)\}$.

\medskip
(i) Prove by induction on $i$ that $Q_{r}^{i}$ is the set of
all states reachable from $q_0$ using paths of length $i$
(where $i$ counts the number of edges).

\medskip
Give an example of a DFA such that
$Q_{r}^{i+1} \not= Q_{r}^{i}$ for all $i\geq 0$.

\medskip
(ii) Give an example of a DFA such that
$Q_{r}^{i} \not= Q_{r}$
for all $i \geq 0$.

\medskip
(iii) Change the inductive definition of $Q^i_r$ as follows:
\[
Q_{r}^{i+1} = Q^i_r\cup
\{q\in Q \mid \exists p\in Q_{r}^{i}, \exists a\in\Sigma,\>
q = \delta(p,a)\}.
\]
Prove that there is a smallest integer $i_0$ such that
\[
Q_{r}^{i_{0}+1} = Q_{r}^{i_{0}} = Q_{r}.
\]

Define the DFA $D_r$ as follows:
$D_r = (Q_r,\Sigma,\delta_r,q_0,F\cap Q_r)$,
where $\mapdef{\delta_r}{Q_r\times\Sigma}{Q_r}$
is the restriction of $\delta$ to $Q_r$.
Explain why $D_r$ is indeed a DFA, and 
prove that $L(D_r) = L(D)$.
A DFA is said to be {\it reachable, or trim\/}, if
$D = D_r$.

% cis511-10hw1
\vspace{0.25cm}
\noindent
{\bf Problem B2 (50 pts).} 
Given any two relatively prime integers $p, q\geq 1$, with $p\not= q$,
($p$ and $q$ are relatively prime iff
their greatest common divisor  is $1$), consider
the language
$L = \{a^p, a^q\}^*$. Prove that
$$\{a^p, a^q\}^* = \{a^n\ |\ n \geq (p-1)(q-1)\}\cup F,$$
where $F$ is some finite set of strings (of length $< (p - 1)(q - 1)$).
Prove that $L$ is a regular language.

\medskip\noindent
{\bf Extra Credit (20 pts).}
Given any two relatively prime integers $p, q\geq 1$, with $p\not= q$,
prove that $pq - p - q = (p-1)(q-1) -1$
is the largest integer not
expressible as $ph + kq$ with $h, k\geq 0$.

\vspace {0.25cm}
\noindent
{\bf Problem B3 (30 pts).} 
Given any alphabet $\Sigma$, prove the following
property: for any two strings $u, v\in\Sigma^*$, $uv = vu$ iff
there is some $w\in\Sigma^*$ such that $u = w^m$ and $v = w^n$, for
some $m, n\geq 0$.



\vspace{0.25cm}
\noindent
{\bf Problem B4 (60 pts).}
Given any two DFA's 
$D_1 = (Q_1, \Sigma, \delta_1, q_{0, 1}, F_1)$
and \\
$D_2 = (Q_2, \Sigma$, $\delta_2, q_{0, 2}, F_2)$,
a {\it morphism $\mapdef{h}{D_1}{D_2}$ of DFA's\/} is a function
$\mapdef{h}{Q_1}{Q_2}$ satisfying the following two conditions:

\begin{enumerate}
\item[(1)]
$h(\delta_1(p, a)) = \delta_2(h(p), a)$,
for all $p\in Q_1$ and all $a\in \Sigma$;
\item[(2)]
$h(q_{0, 1}) = q_{0, 2}$.
\end{enumerate}

An {\it $F$-map $\mapdef{h}{D_1}{D_2}$ of DFA's\/} is a morphism
satisfying the condition 
\begin{enumerate}
\item[(3a)]
$h(F_1) \subseteq F_2$.
\end{enumerate}

\medskip
A {\it $B$-map $\mapdef{h}{D_1}{D_2}$ of DFA's\/} is a morphism
satisfying the condition
\begin{enumerate}
\item[(3b)]
$h^{-1}(F_2) \subseteq F_1$.
\end{enumerate}

\medskip
A {\it proper homomorphism of DFA's\/} is an $F$-map
of DFA's which is also a $B$-map of DFA's, i.e.
it satisfies the condition
\begin{enumerate}
\item[(3c)]
$h^{-1}(F_2) = F_1$.
\end{enumerate}

\medskip
We say that a morphism (resp. $F$-map, resp. $B$-map)
$\mapdef{h}{D_1}{D_2}$  is 
{\it surjective\/} if $h(Q_1) = Q_2$.

\medskip
(a)
Prove that
if $\mapdef{f}{D_1}{D_2}$ and $\mapdef{g}{D_2}{D_3}$ are
morphisms (resp. $F$-maps, resp $B$-maps) of DFAs,
then $\mapdef{g\circ f}{D_1}{D_3}$ is also a morphism
(resp. $F$-map, resp $B$-map) of DFAs.

\medskip
Prove that if  $\mapdef{f}{D_1}{D_2}$ is an $F$-map that is
an isomorphism then it is also a $B$-map, and that 
if  $\mapdef{f}{D_1}{D_2}$ is a $B$-map that is
an isomorphism then it is also an $F$-map.


\medskip
(b)
If $\mapdef{h}{D_1}{D_2}$  is a morphism
of DFA's, prove that
$$h(\delta_1^*(p, w)) = \delta_2^*(h(p), w),$$
for all $p\in Q_1$ and all $w\in \Sigma^*$.

\medskip
As a consequence, prove the following facts:

\medskip
If $\mapdef{h}{D_1}{D_2}$  is an $F$-map of DFA's, then
$L(D_1) \subseteq L(D_2)$.
If $\mapdef{h}{D_1}{D_2}$  is a $B$-map of DFA's, then
$L(D_2) \subseteq L(D_1)$.
Finally, if $\mapdef{h}{D_1}{D_2}$  is a proper  homomorphism of DFA's, then
$L(D_1) = L(D_2)$.

\medskip
(c)
Let $D_1$ and $D_2$ be DFA's and assume that there is a
morphism   $\mapdef{h}{D_1}{D_2}$.
Prove that $h$ induces a unique surjective morphism
$\mapdef{h_r}{(D_1)_r}{(D_2)_r}$
(where $(D_1)_r$ and $(D_2)_r$ are the trim DFA's
defined in problem B1). This means that
if   $\mapdef{h}{D_1}{D_2}$ and   $\mapdef{h'}{D_1}{D_2}$
are DFA morphisms, then $h(p) = h'(p)$ for all
$p\in (Q_1)_r$,
and the restriction of $h$ to $(D_1)_r$ is surjective onto $(D_2)_r$.
Moreover, if $L(D_1) = L(D_2)$,
prove that $h$ induces a unique surjective
proper  homomorphism   $\mapdef{h_r}{(D_1)_r}{(D_2)_r}$.

\medskip
(d)
Relax the condition that a DFA morphism
$\mapdef{h}{D_1}{D_2}$ maps $q_{0, 1}$ to $q_{0, 2}$
(so, it is possible that $h(q_{0, 1})\not= q_{0, 2}$), and call
such a function a {\it weak morphism\/}.
We have an obvious notion of {\it weak $F$-map\/},
{\it weak $B$-map\/} and {\it weak proper
homomorphism\/}
(by imposing condition (3a) or
condition (3b), or (3c)).
For any language, $L \subseteq \Sigma^*$ and any fixed
string, $u\in \Sigma^*$, let $D_u(L)$, also denoted
$L/u$ (called the  {\it (left) derivative of $L$ by $u$\/}), be
the language
\[
D_u(L) = \{v\in \Sigma^*\mid uv \in L\}.
\]
Prove the following facts, 
{\bf assuming that $D_2$ is trim}:
If $\mapdef{h}{D_1}{D_2}$  is a weak $F$-map of DFA's, then
$L(D_1) \subseteq D_u(L(D_2))$, 
for some suitable $u\in\Sigma^*$.
If $\mapdef{h}{D_1}{D_2}$  is a weak $B$-map of DFA's, then
$D_u(L(D_2)) \subseteq L(D_1)$,
for the same $u$ as above.
Finally, if $\mapdef{h}{D_1}{D_2}$  is a weak proper  
homomorphism of DFA's, then
$L(D_1) = D_u(L(D_2))$,
for the same $u$ as above.

\medskip
Suppose there is a weak morphism
$\mapdef{h}{D_1}{D_2}$. What can you say about
the restriction of $h$ to $(D_1)_r$?
What can you say about surjectivity ?
(you may need
to consider $(D_2)_r$ with respect to a {\bf different}
start state). What happens (and what can you say)
if $D_2$ is {\bf not} trim?




\vspace {0.25cm}\noindent
{\bf Problem B5 (50 pts).} ({\it Ultimate periodicity\/})
A subset $U$ of the set $\natnums = \{0, 1, 2, 3,\ldots\}$
of natural numbers is {\it ultimately periodic\/}
if there exist $m, p\in \natnums$, with $p \geq 1$, so that
$n\in U$ iff $n + p\in U$, for all $n\geq m$.

\medskip
(i)
Prove that $U\subseteq \natnums$ is ultimately periodic 
iff either $U$ is finite or there is a finite subset 
$F\subseteq \natnums$
and there are $k \leq p$ 
numbers $m_1, \ldots, m_k$, with
$m_1 < m_2 < \cdots < m_k < m_1 + p$,
and with $m_1$ the smallest
element of $U$ so that for some $p \geq 1$,
$n\in U$ iff $n + p\in U$, for all $n\geq m_1$,
so that 
\[U = F \cup \bigcup_{i = 1}^k \, \{m_i + jp\mid j\in \natnums\}.\]

\medskip
Give an example of an ultimately periodic set $U$
such that $m$ and $p$ are not necessarily unique, i.e.,
$U$ is ultimately periodic with respect to
$m_1, p_1$ and $m_2, p_2$, with $m_1\not= m_2$ and $p_1\not= p_2$.

\remark
A subset of $\natnums$ of the form
$\{m + i p \mid i\in \natnums\}$ (allowing $p = 0$)
is called a {\it linear set\/}, and
a finite union of linear sets is called a {\it semilinear set\/}.
Thus, (i) says that a set is ultimately periodic iff it is semilinear.

\medskip
(ii)
Let $L\subseteq \{a\}^*$ be a language over the one-letter
alphabet $\{a\}$. Prove that $L$ is a regular language
iff the set
$\{m \in \natnums \mid a^m\in L\}$ is ultimately
periodic.
Prove that the family of semilinear sets is closed
under union, intersection and complementation
(i.e., it is a boolean algebra).

\medskip
(iii)
Let $L \subseteq \Sigma^*$ be a regular language over
any alphabet $\Sigma$ (not necessarily
consisting of a single letter).
Prove that the set
\[|L| = \{|w| \mid w \in L\}\]
is ultimately periodic. 


\vspace {0.25cm}\noindent
{\bf Problem B6 (70 pts).} 
(a)
Given any two DFA's $D_1$ and $D_2$, prove that there is a DFA
$D$ and two  DFA $F$-maps $\mapdef{\pi_1}{D}{D_1}$
and $\mapdef{\pi_2}{D}{D_2}$ such that
the following {\it universal mapping property of products\/} holds:
For any DFA $M$ and any two DFA $F$-maps \\
$\mapdef{f}{M}{D_1}$ and $\mapdef{g}{M}{D_2}$, there is
a {\it unique\/} DFA $F$-map $\mapdef{h}{M}{D}$ such that
\[f = \pi_1\circ h
\quad\hbox{and}\quad 
g = \pi_2\circ h,\]
as shown in the diagram below:
%\[\matrice{
%                       &                     &               D_1        \cr
%             & \mapdiagul{f}        & \mapupr{\pi_1}          \cr
%M            & \maprightu{h}                  &  D                    \cr
%             & \mapdiagdr{g}        & \mapdownr{\pi_2}                    \cr
%                       &                     &               D_2        \cr
%}\]
\[
\xymatrix{
  &   \> D_1 \\
M     \ar[ru]^{f} \ar[r]^{h} \ar[rd]_{g} & 
  D   \ar[u]_-{\pi_1}    \ar[d]^-{\pi_2}      \\
  &  \> D_2 
}
\]
Moreover, prove that $\pi_1$ and $\pi_2$ are surjective.
Prove that $D$ is unique up to a DFA $F$-map isomorphism.
This means that if $D'$ is another DFA and if there are two
DFA $F$-maps $\mapdef{\pi_1'}{D'}{D_1}$
and $\mapdef{\pi_2'}{D'}{D_2}$ such that
the universal mapping property of products holds, then
there are two unique DFA $F$-maps $\mapdef{\varphi}{D}{D'}$ and
$\mapdef{\varphi'}{D'}{D}$ so that
$\varphi'\circ \varphi = \id_{D}$,
$\pi_1 = \pi_1'\circ  \varphi$, $\pi_2 =  \pi_2'\circ \varphi$,
$\varphi\circ \varphi' = \id_{D'}$,
$\pi_1' = \pi_1\circ  \varphi'$ and $\pi_2' =  \pi_2\circ \varphi'$.
What is the language accepted by $D$?

\remark
We call $D$ the {\it product of $D_1$ and $D_2$\/}
and we denote it by $D_1\prod D_2$.

\medskip
(b)
Given any three DFA's $D_1$, $D_2$, and $D_3$ 
and any two DFA $F$-maps
$\mapdef{f}{D_1}{D_3}$ and $\mapdef{g}{D_2}{D_3}$, 
prove that there is a DFA
$D$ and two  DFA $F$-maps $\mapdef{\pi_1}{D}{D_1}$
and $\mapdef{\pi_2}{D}{D_2}$ such that
\[f\circ \pi_1 = g\circ \pi_2,\]
as in the diagram below
%\[
%\matrice{
%D            &  \maprightu{\pi_1}  &  D_1          \cr
%\mapdownl{\pi_2}   &                 & \mapdownr{f}\cr
%D_2            & \maprightd{g}&  D_3,            \cr 
%}
%\]
%
\[
\xymatrix{
D \ar[r]^{\pi_1} \ar[d]_{\pi_2} & \> D_1 \ar[d]^{f} \\
D_2 \ar[r]_{g} & \> D_3
}
\]
and the following {\it universal mapping property of 
fibred products\/} holds:
for any DFA $M$ and any two DFA $F$-maps
$\mapdef{f'}{M}{D_1}$ and $\mapdef{g'}{M}{D_2}$
such that
\[f\circ f' = g\circ g',\]
as in the diagram below
%\[
%\matrice{
%M            &  \maprightu{f'}  &  D_1          \cr
%\mapdownl{g'}   &                 & \mapdownr{f}\cr
%D_2            & \maprightd{g}&  D_3,            \cr 
%}
%\]
%
\[
\xymatrix{
M \ar[r]^{f'} \ar[d]_{g'} & \> D_1 \ar[d]^{f} \\
D_2 \ar[r]_{g} & \> D_3
}
\]
there is a
{\it unique\/} DFA $F$-map $\mapdef{h}{M}{D}$ such that
\[f' = \pi_1\circ h
\quad\hbox{and}\quad 
g' = \pi_2\circ h,\]
as in the diagram below
\[
\xymatrix{
M   \> \ar[rrd]^-{f'} \ar[ddr]_-{g'} \ar@{.>}[rd] |-{h} &  & \> \\
&  D \> \ar[r]_-{\pi_1} \ar[d]^-{\pi_2} & \> D_1 \ar[d]^{f} \\
&  D_2\> \ar[r]_{g} & \> D_3
}
\]
Prove that $D$ is unique up to a DFA $F$-map isomorphism.
This means that if $D'$ is another DFA and if there are
two DFA $F$-maps  $\mapdef{\pi_1'}{D'}{D_1}$
and $\mapdef{\pi_2'}{D'}{D_2}$ such that
\[f\circ \pi_1' = g\circ \pi_2'\]
and the universal mapping property of 
fibred products holds, then
there are two unique DFA $F$-maps $\mapdef{\varphi}{D}{D'}$ and
$\mapdef{\varphi'}{D'}{D}$ so that
$\varphi'\circ \varphi = \id_{D}$
$\pi_1 = \pi_1'\circ  \varphi$, $\pi_2 =  \pi_2'\circ \varphi$,
$\varphi\circ \varphi' = \id_{D'}$,
$\pi_1' = \pi_1\circ  \varphi'$ and $\pi_2' =  \pi_2\circ \varphi'$.


\remark
We denote $D$ by $D_1\prod_{D_3} D_2$ and call it
a {\it fibred product of $D_1$ and $D_2$ over $D_3$\/},
or a {\it pullback of $D_1$ and $D_2$ over $D_3$\/}.

\medskip
If  $T$ is any one-state DFA accepting $\Sigma^*$
(this single state is accepting), observe that there is a unique
DFA $F$-map from every DFA $D$ to $T$.
Use this to show that if $D_1\prod D_2$ is the
product DFA arising in (a), then 
\[D_1\prod D_2 = D_1\prod_{T} D_2.\]


\vspace{0.25cm}\noindent
{\bf Extra Credit (40 points).}
Redo questions (a) and (b) for $B$-maps instead of
$F$-maps. 

\remark
If we dualize (b), i.e.,
turn the arrows around, we get the notion of
{\it fibred coproduct\/} or {\it pushout\/}.
It can be shown that fibred coproducts exist,
both for $F$-maps and $B$-maps,
but this is  tricky.

\vspace{0.5cm}\noindent
{\bf TOTAL: 290 + 60 points.}

\end{document}



