\documentclass[12pt]{article}
\usepackage{fullpage}
\usepackage{titlesec}
\usepackage{tikz}
\usepackage{amsfonts,amssymb}
\usepackage{amsmath}
\usepackage{comment}
\usetikzlibrary{automata, positioning}

\input ../libraries/mac.tex
\input ../libraries/mathmac.tex

\begin{document}
\pagestyle{plain}
\titleformat{\subsection}[runin]
  {\normalfont\large\bfseries}{\thesubsection}{1em}{}

\title{Homework 5}
\author{Brooke Fugate, Michael O'Connor, Rohan Shah}
\date{}

\maketitle

\section*{Problem B2}
\subsection*{(i)} Prove that the language $L_1 = \{a^mb^nc^p \mid 1\le m<n<p\}$
is not context-free
\medskip \newline
Proof by contradiction, so assume $L_1$ is context-free, then it must also
satisfy Ogden's lemma. Let $m$ be the constant $K > 1$ for Ogden's lemma.
Let $w = a^mb^{m+1}c^{m+2}$ such that $w \in L_1$. Mark every $a$ in $w$,
no $b$'s or $c$'s are marked. The condition that $w$ contains at least $m$
marked occurences holds since it contains exactly $m$ marked occurences.
By Ogden's lemma there exists some decomposition of $w = uvxyz$. By applying
each of the properties of Ogden's lemma (except the pumping property) we can
narrow down the decomposition of $w$ with the following restrictions:
\begin{itemize}
\item $u=a^i$ such that $i\ge 0$
\item $v=a^i$ such that $i\ge 0$
\item $x=a^ib^jc^k$ such that $i\ge 1$ and $j,k\ge 0$
\item $y=a^ib^jc^k$ such that $i,j,k\ge 0$
\item $z=a^ib^jc^k$ such that $i,j,k\ge 0$
\item $u=a^i,\ v=a^j$ such that $i,j\ge 1$ \textbf{or}
$y=a^i,\ z=a^jb^{m+1}c^{m+2}$ such that $i,j\ge 1$
\item $vxy = a^ib^jc^k$ such that $i<m$ and $j,k\ge 0$
\end{itemize}
Now we can further narrow down the decomposition of $w$ by applying the pumping
property, that is, since $w = uvxyz \in L(G)$ then $uv^nxy^nz \in L(G)$ for all
$n \ge 0$. By this property we can further restrict $y$ to the form
\begin{itemize}
\item $y=c^k$ such that $k\ge 0$
\end{itemize}
since pumping any other $y \in \Sigma^*$ would break the pumping property. We
can easily prove this by case analysis. If $y$ contained more than one symbol
from $\Sigma$, i.e. $y=a^ib^jc^k$, $y=a^ib^j$ or $y=b^jc^k$ such that $i,j,k\ge
1$, then pumping it would cause the new string to contain a substring of the
form $c^ka^i$, $b^ja^i$, or $c^kb^j$ such that $i,j,k\ge 1$ respectively,
which means that the pumped string is not in $L_1$. Thus $y$ must only contain
a single symbol from $\Sigma$. But if $y=a^i$ or $y=b^j$ such that $i,j\ge 1$
then it would again break the pumping property since it implies that $z$ must
contain all the $p$ $c$'s in $w$ and thus pumping $y$ would eventually cause
the pumped string to contain more $a$'s or $b$'s then $c$'s respectively,
which again means the pumped string is not in $L_1$. Thus we have the above
restriction to $y$ i.e. that $y=c^k$ such that $k\ge 0$. With this restriction
and we can again use the properties of Ogden's lemma to restrict $u$, $v$, and
$x$ to the forms:
\begin{itemize}
\item $x=a^ib^nc^k$ such that $i\ge 1$ and $k\ge 0$
\item $u=a^i,\ v=a^j$ such that $i,j\ge 1$
\end{itemize}
since $x$ must contain a marked occurence and therefore must also contain all
$b$'s in $w$ since $y$ contains only $c$'s. Further, $u$ and $v$ must contain
marked occurences since it is impossible for both $y$ and $z$ to. This final
form of the decomposition of $w=uvxyz$ also breaks the pumping property since
pumping $v$ will cause the number of $a$'s to be greater than the number of
$b$'s (which don't get pumped since they are all in $x$) and thus the pumped
string is not in $L_1$. Therefore, Ogden's lemma does not hold for the language
$L_1$, which is a contradiction, so $L_1$ is not context-free.

\subsection*{(ii)}Prove that the language
$L_2 = \{a^nb^nc^p \mid n,p\ge 1, p\neq n\}$ is not context-free.
\medskip \newline
Proof by contradiction, so assume $L_2$ is context-free, then it must also
satisfy Ogden's lemma. Let $m$ be the constant $K > 1$ for Ogden's lemma.
Let $p=m!$. Let $w = a^{2p}b^{2p}c^p$ such that $w \in L_2$. Mark every $c$ in
$w$, no $a$'s or $b$'s are marked. The condition that $w$ contains at least
$m$ marked occurences holds since it contains $m!$ marked occurences.
By Ogden's lemma there exists some decomposition of $w = uvxyz$ of the following
form:
\begin{itemize}
\item $u=a^ib^jc^k$ such that $i,j,k\ge 0$
\item $v=a^ib^jc^k$ such that $i,j,k\ge 0$
\item $x=a^ib^jc^k$ such that $i,j\ge 0$ and $k\ge 1$
\item $y=c^i$ such that $i\ge 0$
\item $z=c^i$ such that $i\ge 0$
\item $u=a^{2p}b^{2p}c^i,\ v=c^j$ such that $i,j\ge 1$ \textbf{or}
$y=c^i,\ z=c^j$ such that $i,j\ge 1$
\item $vxy = a^ib^{2p}c^k$ such that $0\le k<m$
\end{itemize}
Now we can further narrow down the decomposition of $w$ by applying the pumping
property, that is, since $w = uvxyz \in L(G)$ then $uv^nxy^nz \in L(G)$ for all
$n \ge 0$. By this property we can further restrict $v$ to the form
\begin{itemize}
\item $v=c^k$ such that $k\ge 0$
\end{itemize}
since pumping any other $v \in \Sigma^*$ would break the pumping property. We
can easily prove this by case analysis. If $v$ contained more than one symbol
from $\Sigma$, i.e. $v=a^ib^jc^k$, $v=a^ib^j$ or $v=b^jc^k$ such that $i,j,k\ge
1$, then pumping it would cause the new string to contain a substring of the
form $c^ka^i$, $b^ja^i$, or $c^kb^j$ such that $i,j,k\ge 1$ respectively,
which means that the pumped string is not in $L_2$. Thus $y$ must only contain
a single symbol from $\Sigma$. But if $y=a^i$ such that $i\ge 1$ then it would
break the pumping property since it implies that $x$ would contain all the $2p$
$b$'s in $w$ in which case pumping $v$ would cause the pumped string to contain
more $a$'s than $b$'s which means the pumped string would not be in the language.
And if $y=b^j$ such that $j\ge 1$ then it would again break the pumping property
since it implies that $u$ would contain all the $2p$ $a$'s in $w$ in which case
pumping $v$ would cause the pumped string to contain more $b$'s than $a$'s which
means the pumped string would not be in the language. Thus we have the above
restriction to $v$ i.e. that $v=c^k$ such that $k\ge 0$. With this we can
further restrict the decomposition of $w$ to the following form:
$$u=a^{2p}b^{2p}c^f, v=c^h, x=c^j, y=c^k,\text{ and } z=c^l
\text{ such that } f,h,k,l\ge 0 \text{ and } j\ge1$$
And so we have that:
$$w = a^{2p}b^{2p}c^{f+h+j+k+l} \text{ where } f,h,k,l\ge 0, j\ge 1,
\text{ and } f+h+j+k+l=p$$
which by the pumping property of Ogden's lemma implies that
$$w = a^{2p}b^{2p}c^{i(h+k) + (f+j+l)} \text { where } i(h+k)+(f+j+l)\neq 2p
\text{ for all } i\ge 0$$
where either $f,h\ge 1$ or $k,l\ge 1$ by the property of Ogden's lemma that
either both $u$ and $v$ or both $y$ and $z$ must contain some marked occurences.
So, for Ogden's lemma to hold it must be the case that:
$$i \neq \frac{2p-(f+j+l)}{(h+k)} \text{ for all } i\ge 0$$
where $f,h,k,l \ge 0, j\ge 1,$ either $f,h\ge 1$ or $k,l\ge 1$ and
$f+j+l+h+l = p$ i.e. $(f+j+l) \le p \le 2p$ and $(h+k) \le p \le 2p$. And we
know that
$$p = m!$$
where $h+j+k < m$ by the property of Ogden's lemma that $v,x,y$ must contain
less than $m$ marked occurences. Which means that
$$\frac{2p}{(h+k)}-\frac{(f+j+l)}{(h+k)}$$
is a positive integer which is a contradiction since $i\ge 0$ is any positive
integer therefore Ogden's lemma does not hold, so $L_2$ is not context-free.

\end{document}
